Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/VIRGO Network

### 2013-05-21 doc/paper.tex
added diegos comments

\newcommand{\carl}[1]{{\color{red}  #1}}
\newcommand{\ilya}[1]{{\color{blue} #1}}
\newcommand{\ben}[1]{{\color{green} #1}}
\newcommand{\will}[1]{{\color{cyan} #1}}

@@

 By 2016, the field of gravitational-wave astrophysics will
 mature from a speculative scientific genre into an ideal tool for 
 exploring the gravitational side of the universe. Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \carl{[CITE]}.
 The Advanced LIGO and VIRGO detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense preparations are currently
 underway to prepare to characterize and extract as much physical 
 information as possible from these signals.

@@

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream to a bank of theoretical templates, the roughly 16KHz
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection is claimed.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to near identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique 
and proper sampling, we will be able to make precise, scientifically meaningful
statements about the physics of these detectible systemss.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/VIRGO era, our results give the first estimate at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among 
many others) have done their due diligence in pointing out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which states
the BNS parameter estimation capabilities of Advanced LIGO/VIRGO via the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations about certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in our gravitational wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with mean zero.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as the probability
that our signal is Gaussian once the waveform has been subtracted
\begin{align}
  p(s | \boldsymbol{\theta}) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\boldsymbol{\theta}) | s-h(\boldsymbol{\theta})\right >\right] ,
  \label{likelihood}
\end{align}
where $\boldsymbol{\theta}$ is the set of parameters for our template
waveforms.  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

Once we have the likelihood of the signal \eqref{likelihood}, we
employ Bayes Rule to obtain the posterior probability of the system
parameters $\boldsymbol{\theta}$ given the signal $s$ as
\begin{align}
  p(\boldsymbol{\theta} | s) &= \frac{p(\boldsymbol{\theta})p(s | \boldsymbol{\theta})}{p(s)} \nonumber\\
  & \propto p(\boldsymbol{\theta}) \exp\left[-\frac{1}{2}\big < s - h(\boldsymbol{\theta}) | s-h(\boldsymbol{\theta}) \big > \right] ,
  \label{posterior}
\end{align}
where $p(\boldsymbol{\theta})$ are the prior probabilities on our
source parameters and $p(s)$ is a normalization constant.  Our prior
information can come from physics limits in the parameter space, or
from our \textit{a priori} knowledge of astrophysical systems. If we
pick a set of parameters $\boldsymbol{\theta}$ such that
$h(\boldsymbol{\theta}) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\boldsymbol{\theta}_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\boldsymbol{\theta}|s\right)$.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.  Both samplers were tuned and 
developed during the last science run of the Initial LIGO/VIRGO network.  
A description of the parameter estimation capabilities, c.f. 2010, can be found
in \cite{S6PE}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = A f^{-7/6}e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $A \propto \mathcal{M}_c^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M\equiv M_1+M_2$, it is convenient to work with the
\textit{mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\mathcal{M}_c = (M_1 M_2)^{3/5} M^{-1/5},
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead use the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is empirically superior when proposing MCMC parameter jumps, and is more physically intuitive.}  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M}{\mathcal{M}_c}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

@@

The terms $t_c$ and $\phi_0$ in equation (\ref{phase}) are constants
of integration, referring to the chirp time and coalescence phase,
respectively.  Although uninteresting physically, they must be
accounted for in any parameter estimation study of the waveform phase.

To perform the integral defined in \eqref{innerProduct}, we used the best estimate
for a high-power, zero-detuning configuration of Advanced LIGO, provided by the LIGO Scientific Collaboration.
Both the noise curve and technical reports describing it can be found in \cite{ADVLIGONoise}.
We consider two configurations of the advanced detector
network: a three-detector configuation consistanting of the two LIGO sites 
(in Hanford, WA and Livingston, LA) and the Virgo site (in Pisa), and a four-detector
configuration with adds the proposed LIGO-India detector (in Chitradurga, KA).
For simplicity, we assume each detector to be operating at the Advanced LIGO sensitivity. 
For a multi-detector network, the noise-weighted inner products
\eqref{innerProduct} combine linearly, allowing us to use the above
formalism with minimal modification.  We integrate the inner product
from a lower-frequency cutoff of $20\text{Hz}$ to the
innermost-stable-circular orbit of the systems in question, which for a
non-spinning binary is a function only of the total mass:

@@

We consier the mass parameters ($\mathcal{M}_c,q$) and the phasing parameters,
($\phi_0,t_c$), to be the intrinsic parameters of a gravitational-wave signal.  There 
are an additional 5 extrinsic parameters, independent of the pN orbital phase, which can
modulate the amplitude of the signal in an independent fashion for each detector.  
Combining these leads to our 9-dimensional parameter space for non-spinning systems:
\begin{equation}
\boldsymbol{\theta} =  (\mathcal{M}_c, q, \phi_0,t_c,D,\iota,\psi,\alpha,\delta)
\label{parameterspace}
\end{equation}
where to summarize

\begin{itemize}
\item $\mathcal{M}_c$ is the chirp mass,
\item $q$ is the mass ratio,
\item $\phi_0$ and $t_c$ are the chirp phase and chirp time, arbitrary phasing parameters,
\item $D$ is the luminosity distance to the binary,
\item $\iota$ is the orbital inclination, 
\item $\psi$ the gravitational-wave polarization, and
\item $\alpha$ and $\delta$ the right ascension and declination of the
source on the sky.
\end{itemize}
  Since the wave amplitude depends on the orientation of the
binary with respect to each detector, most of the information about these extrinsic
parameters comes from two sources: the time-of-arrival triangulation of 
the source locations, and the relative amplitudes within the detector network.

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector is
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}^{*}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

@@

In Figure~\ref{9dPDF}, we provide an example of the 9, 1-dimensional marginalized posterior 
probability density functions recovered from a single $1.4M_{\odot}/1.4M_{\odot}$ BNS system. 
 These PDFs are typical of what will be produced by parameter estimation studies in the advanced 
 detector era.  Notice that the peak of several parameters, including the chirp mass, $\mathcal{M}_c$, 
 appears to be displaced from the true values in red.  This effect is due to the reduction of the
  9-dimensional PDF to a series of marginalized 1-dimensional PDFs.  For instance, the 1-dimensional 
  PDF for chirp mass is marginalized via

\begin{equation}
p(\mathcal{M}_c | s) = \int _{\boldsymbol{\theta} \setminus \mathcal{M}_c} p(\boldsymbol{\theta} | s)~d(\boldsymbol{\theta} \setminus \mathcal{M}_c)
\label{mcMarginalization}
\end{equation}

\noindent Other parameter and higher-dimensional marginalizations follow a similar convention.

@@

\begin{figure}[h!]
  \centering
 \includegraphics[trim=0cm 0cm 2cm 0cm, clip=false,scale=0.56]{1414masses2D.pdf}
 \caption{2D marginalized posterior probability density functions for the mass parameters recovered in typical a $1.4M_{\odot}/1.4M_{\odot}$ system.  The posteriors are plotted in terms of parameters used in the waveform, chirp mass ($\mathcal{M}_C$) and the mass ratio ($q$), and in the individual component masses of the binary ($M_1$ and $M_2$).  In the $\mathcal{M}_c$-$q$ space, the posterior would resemble a Gaussian if not for the limitation of $q \leq 1$.  The presence of the $q$ cutoff and the convention that $M_1 \geq M_2$ inform the non-Gaussian features present.  When projected as 1D marginalized posteriors, the component masses resemble the posterior PDFs shown in figure \ref{metaMassPDFs}.}
  \label{1414masses}
\end{figure}

\begin{figure}[h!]
  \centering
 \includegraphics[trim=0cm 0cm 2cm 0cm, clip=false,scale=0.56]{125masses2D.pdf}
 \caption{Similar to Figure~\ref{1414masses}, for a typical $1M_{\odot}/2.5M_{\odot}$ system.  The unequal mass ratio displaces the posterior PDFs from the $q \leq 1$ present in the equal mass case, yielding a Gaussian PDF in both the $\mathcal{M}_c$-$q$ and $M_1$-$M_2$ spaces.}
  \label{125masses}
\end{figure}

@@

Of the 9 variables in our parameter space \eqref{parameterspace}, the two of immediate physical
interest are the two parameters, $\mathcal{M}_c$ and $q$, or correspondingly the direct masses 
of the two companions, $M_1$ and $M_2$.    As stated above, the ability of
Advanced LIGO/VIRGO to construct a population of BNS masses will be critical to the science
potential of gravitational-wave astronomy.

@@

In practice, it was found that the cross-correlations between the extrinsic parameters and signal time-of-arrival were several orders of magnitude \carl{[QUANTIFY]} below the uncertainties in $\mathcal{M}_c$ and $q$.  It should be noted that, in practice, this will \emph{not} be the case: in reality, the noise power-spectral density of each detector varies as a function of time, including both natural variations in sensitivity and detector glitches \carl{[CITE]}.  In such a case, the base sensitivity of a gravitational-wave detector to even the easiest to recover parameters will vary as a function of position and time.

In Figures~\ref{1414masses} and~\ref{125masses}, we show the marginalized 2D posterior PDFs of our mass parameters a prototypical equal mass and unequal mass binaries.  We include the PDF in both the $\mathcal{M}_c$-$q$ space (relevant for the waveform and the MCMC algorithm), and the more physically interesting component mass space ($M_1$-$M_2$).  Although only the $1.4M_{\odot}/1.4M_{\odot}$ system is included in Figure \ref{1414masses}, the PDF is virtually identical to the other equal mass cases, modulo a scaling factor.  Again, this is unsurprising due to the identical SNR of each of the injected signals.

@@

Since it is assumed that binary neutron star systems (along with neutron star/black systems) are the progenitors for short gamma-ray bursts, emitting the burst along the axis of orbital angular momentum, the inclination of the binary system will be of particular interest to gamma-ray astronomy \carl{[CITE]}.  The inclination is detected as a relative amplitude difference between the two gravitational-wave polarization, such that to lowest order

@@

\noindent  yields a detection.  This requirement provides us with a convenient test for the effectiveness 
of our parameter estimation.  First, we note that maximizing our likelihood function, \eqref{likelihood},
is identical to maximizing the SNR \eqref{formalSNR} of our template waveform in a given stretch of data.
Therefore, the better the match between the template waveform and the data, the higher the SNR.  Since we
assume, quite tautologically, that our parameter estimation code should estimate parameters at least as 
well as a matched filtering approach, we would be immediately suspicious of any MCMC run whose 
maximum-likelihood waveform had an SNR below that of our pipeline trigger.   In that case, we would perform
the MCMC with a much longer runtime, until the maximum likelihood of our posterior distribution had an SNR
at least as high as the detection.

@@

\begin{figure}[ht!]
  \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, clip=true,scale=0.59]{noisePDF.pdf}
 \label{noisePDFs}
 \caption{The effects of a non-zero noise realization on the recovered PDF for five different Gaussian realizations of the Advanced LIGO noise curve.  Each PDF represents the recovery of the same $1M_{\odot}/2.5M_{\odot}$ signal in a different noise curve, picked at random from the Gaussian colored noise defined by the Advanced LIGO power spectral density.  Notice how the curve is relatively unchanged, but usually assumes a Gaussian PDF displaced from the true value.  This only holds true for glitch-free data, which is an unrealistic idealization when compared to real data.}
\end{figure}

@@

As a demonstration of this, we injected a single $1.4M_{\odot}/1.4M_{\odot}$ system, 
detected in HLVI,  into 5 separate realizations of Gaussian noise colored by the Advanced LIGO PSD, 
and compared the results to the same MCMC recovery in the zero-noise case.  This example
can be seen for one-dimensional marginalization of $\mathcal{M}_c$ and $q$ in Fig. \ref{noisePDFs}.  In effect, the ``real answer'' that will be recovered is a single PDF with similar 
width to the zero-noise PDF, but with the peak likelihood displaced from the true value.

Of course, the largest concern for gravitational-wave parameter estimation will not the specific
realization of Gaussian noise, but the effects of glitches in the data.  These time-series 
glitches have a variety of well-known and not-so-well-known physical causes in the instrument,
and can cause substantially greater errors in parameter recovery than any specific realization noise realizations.  This has already been addressed for the case of space-based gravitational-wave
detectors (LISA), and is currently being implemented in \texttt{lalinference\_mcmc} 
\carl{[CITE TYSON, PROTO-TYSON]}.

>>>>

\newcommand{\carl}[1]{{\color{red}  #1}}
\newcommand{\tyson}[1]{{\color{blue} #1}}
\newcommand{\ben}[1]{{\color{green} #1}}
\newcommand{\will}[1]{{\color{cyan} #1}}
\newcommand{\vivien}[1]{{\color{purple} #1}}
\newcommand{\diego}[1]{{\color{pink} #1}}

\newcommand{\thpara}{\boldsymbol{\theta}}
\newcommand{\chmass}{\mathcal{M}_c}

@@

 By 2016, the field of gravitational-wave astrophysics will
 mature into an ideal tool for 
 exploring the gravitational side of the universe. Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \carl{[CITE]}.
 The Advanced LIGO and VIRGO detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense work is being undertaken
 to prepare to characterize and extract as much physical 
 information as possible from these signals.

@@

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream with a bank of theoretical templates, the roughly 16KHz
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection candidate is found.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to nearly identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique 
and proper sampling, we will be able to make precise, scientifically meaningful
statements about the physics of these detectible systems.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/VIRGO era, our results give the first estimate at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among 
many others) have pointed out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which investigates
the BNS parameter estimation capabilities of Advanced LIGO/VIRGO via the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations of certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with mean zero.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as proportional to the probability
that our signal is Gaussian once the waveform has been subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for our template
waveforms.  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

Once we have the likelihood of the signal \eqref{likelihood}, we
employ Bayes Rule to obtain the posterior probability of the system
parameters $\thpara$ given the signal $s$ as
\begin{align}
  p(\thpara | s) &= \frac{p(\thpara)p(s | \thpara)}{p(s)} \nonumber\\
  & \propto p(\thpara) \exp\left[-\frac{1}{2}\big < s - h(\thpara) | s-h(\thpara) \big > \right] ,
  \label{posterior}
\end{align}
where $p(\thpara)$ are the prior probabilities on our
source parameters and $p(s)$ is a normalization constant.  Our prior
information can come from physics limits in the parameter space, or
from our \textit{a priori} knowledge of astrophysical systems. If we
pick a set of parameters $\thpara$ such that
$h(\thpara) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\thpara_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\thpara|s\right)$.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.  Both samplers were tuned and 
developed during the last science run of the Initial LIGO/VIRGO network.  
A description of the parameter estimation capabilities, c.f. 2010, can be found
in \cite{S6PE}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = A f^{-7/6}e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $A \propto \chmass^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M\equiv M_1+M_2$, it is convenient to work with the
\textit{mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5} M^{-1/5},
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is empirically superior when proposing MCMC parameter jumps, and is more physically intuitive.}  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

@@

The terms $t_c$ and $\phi_0$ in equation (\ref{phase}) are constants
of integration, referred to as the chirp time and coalescence phase,
respectively.  Although uninteresting physically, they must be
accounted for in any parameter estimation study of the waveform phase.

To perform the integral defined in \eqref{innerProduct}, we used as our power-spectral
density the best estimate for a high-power, zero-detuning configuration of Advanced 
LIGO, provided by the LIGO Scientific Collaboration.  Both the noise curve and technical 
reports describing it can be found in \cite{ADVLIGONoise}.
We consider two configurations of the advanced detector
network: a three-detector configuation consistanting of the two LIGO sites 
(in Hanford, WA and Livingston, LA) and the Virgo site (in Pisa, Italy), and a 
four-detector configuration with adds the proposed LIGO-India detector
(in Chitradurga, KA).  For simplicity, we assume each detector to be 
operating at the Advanced LIGO sensitivity.

For a multi-detector network, the noise-weighted inner products
\eqref{innerProduct} combine linearly, allowing us to use the above
formalism with minimal modification.  We integrate the inner product
from a lower-frequency cutoff of $20\text{Hz}$ to the
innermost-stable-circular orbit of the systems in question, which for a
non-spinning binary is a function only of the total mass:

@@

We consier the mass parameters ($\chmass,q$) and the phasing parameters,
($\phi_0,t_c$), to be the intrinsic parameters of a gravitational-wave signal.  There 
are an additional 5 extrinsic parameters, independent of the pN orbital phase, which can
modulate the amplitude of the signal in an independent fashion for each detector.  
Combining these leads to our 9-dimensional parameter space for non-spinning systems:
\begin{equation}
\thpara =  (\chmass, q, \phi_0,t_c,D,\iota,\psi,\alpha,\delta)
\label{parameterspace}
\end{equation}
where to summarize

\begin{itemize}
\item $\chmass$ is the chirp mass,
\item $q$ is the mass ratio,
\item $\phi_0$ and $t_c$ are the chirp phase and chirp time, arbitrary phasing parameters,
\item $D$ is the luminosity distance to the binary,
\item $\iota$ is the orbital inclination, 
\item $\psi$ the gravitational-wave polarization, and
\item $\alpha$ and $\delta$ the right ascension and declination of the
source on the sky.
\end{itemize}
  Since the wave amplitude depends on the orientation of the
binary with respect to each detector, most of the information about these extrinsic
parameters comes from two sources: the time-of-arrival triangulation of 
the source locations, and the relative amplitudes within the detector network.

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector as 
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}^{*}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

@@

In Figure~\ref{9dPDF}, we provide an example of the 9, 1-dimensional marginalized posterior 
probability density functions recovered from a single $1.4M_{\odot}/1.4M_{\odot}$ BNS system. 
 These PDFs are typical of what will be produced by parameter estimation studies in the advanced 
 detector era.  Notice that the peak of several parameters, including the chirp mass, $\chmass$, 
 appears to be displaced from the true values in red.  This effect is due to the reduction of the
  9-dimensional PDF to a series of marginalized 1-dimensional PDFs.  For instance, the 1-dimensional 
  PDF for chirp mass is marginalized via

\begin{equation}
p(\chmass | s) = \int _{\thpara \setminus \chmass} p(\thpara | s)~d(\thpara \setminus \chmass)
\label{mcMarginalization}
\end{equation}

\noindent where the notation $\thpara \setminus \chmass$ implies all
parameter of \eqref{parameterspace} except $\chmass$.  Other parameter and higher-dimensional marginalizations follow a similar convention.

@@

\begin{figure}[h!]
  \centering
 \includegraphics[trim=0cm 0cm 2cm 0cm, clip=false,scale=0.56]{1414masses2D.pdf}
 \caption{2D marginalized posterior probability density functions for the mass parameters recovered in typical a $1.4M_{\odot}/1.4M_{\odot}$ system.  The posteriors are plotted in terms of parameters used in the waveform, chirp mass ($\mathcal{M}_C$) and the mass ratio ($q$), and in the individual component masses of the binary ($M_1$ and $M_2$).  In the $\chmass$-$q$ space, the posterior would resemble a Gaussian if not for the limitation of $q \leq 1$.  The presence of the $q$ cutoff and the convention that $M_1 \geq M_2$ inform the non-Gaussian features present.  When projected as 1D marginalized posteriors, the component masses resemble the posterior PDFs shown in figure \ref{metaMassPDFs}.}
  \label{1414masses}
\end{figure}

\begin{figure}[h!]
  \centering
 \includegraphics[trim=0cm 0cm 2cm 0cm, clip=false,scale=0.56]{125masses2D.pdf}
 \caption{Similar to Figure~\ref{1414masses}, for a typical $1M_{\odot}/2.5M_{\odot}$ system.  The unequal mass ratio displaces the posterior PDFs from the $q \leq 1$ present in the equal mass case, yielding a Gaussian PDF in both the $\chmass$-$q$ and $M_1$-$M_2$ spaces.}
  \label{125masses}
\end{figure}

@@

Of the 9 variables in our parameter space \eqref{parameterspace}, the two of immediate physical
interest are the two parameters, $\chmass$ and $q$, or correspondingly the direct masses 
of the two companions, $M_1$ and $M_2$.    As stated above, the ability of
Advanced LIGO/VIRGO to construct a population of BNS masses will be critical to the science
potential of gravitational-wave astronomy.

@@

In practice, it was found that the cross-correlations between the extrinsic parameters and signal time-of-arrival were several orders of magnitude \carl{[QUANTIFY]} below the uncertainties in $\chmass$ and $q$.  It should be noted that, in practice, this will \emph{not} be the case: in reality, the noise power-spectral density of each detector varies as a function of time, including both natural variations in sensitivity and detector glitches \carl{[CITE]}.  In such a case, the base sensitivity of a gravitational-wave detector to even the easiest to recover parameters will vary as a function of position and time.

In Figures~\ref{1414masses} and~\ref{125masses}, we show the marginalized 2D posterior PDFs of our mass parameters a prototypical equal mass and unequal mass binaries.  We include the PDF in both the $\chmass$-$q$ space (relevant for the waveform and the MCMC algorithm), and the more physically interesting component mass space ($M_1$-$M_2$).  Although only the $1.4M_{\odot}/1.4M_{\odot}$ system is included in Figure \ref{1414masses}, the PDF is virtually identical to the other equal mass cases, modulo a scaling factor.  Again, this is unsurprising due to the identical SNR of each of the injected signals.

@@

Since it is assumed that binary neutron star systems (along with neutron star/black hole systems) are the progenitors for short gamma-ray bursts, emitting the burst along the axis of orbital angular momentum, the inclination of the binary system will be of particular interest to gamma-ray astronomy \carl{[CITE]}.  The inclination is detected as a relative amplitude difference between the two gravitational-wave polarization, such that to lowest order

@@

\noindent  yields a detection candidate.  This requirement provides us with a convenient test for the effectiveness 
of our parameter estimation.  First, we note that maximizing our likelihood function, \eqref{likelihood},
is identical to maximizing the SNR \eqref{formalSNR} of our template waveform in a given stretch of data.
Therefore, the better the match between the template waveform and the data, the higher the SNR.  Since we
assume, quite tautologically, that our parameter estimation code should estimate parameters at least as 
well as a matched filtering approach, we would be immediately suspicious of any MCMC run whose 
maximum-likelihood waveform had an SNR below that of our pipeline trigger.   In that case, we would perform
the MCMC with a much longer runtime, until the maximum likelihood of our posterior distribution had an SNR
at least as high as the detection.

@@

\begin{figure}[ht!]
  \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, clip=true,scale=0.59]{noisePDF.pdf}
 \caption{The effects of a non-zero noise realization on the recovered PDF for five different Gaussian realizations of the Advanced LIGO noise curve.  Each PDF represents the recovery of the same $1M_{\odot}/2.5M_{\odot}$ signal in a different noise curve, picked at random from the Gaussian colored noise defined by the Advanced LIGO power spectral density.  Notice how the curve is relatively unchanged, but usually assumes a Gaussian PDF displaced from the true value.  This only holds true for glitch-free data, which is an unrealistic idealization when compared to real data.}
 \label{noisePDFs}
\end{figure}

@@

As a demonstration of this, we injected a single $1.4M_{\odot}/1.4M_{\odot}$ system, 
detected in HLVI,  into 5 separate realizations of Gaussian noise colored by the Advanced LIGO PSD, 
and compared the results to the same MCMC recovery in the zero-noise case.  This example
can be seen for one-dimensional marginalization of $\chmass$ and $q$ in Fig. \ref{noisePDFs}.  In effect, the ``real answer'' that will be recovered is a single PDF with similar 
width to the zero-noise PDF, but with the peak likelihood displaced from the true value.

Of course, the largest concern for gravitational-wave parameter estimation will not be the specific
realization of Gaussian noise, but the effects of glitches in the data.  These time-series 
glitches have a variety of well-known and not-so-well-known physical causes in the instrument,
and can cause substantially greater errors in parameter recovery than any specific realization noise realizations.  This has already been addressed for the case of space-based gravitational-wave
detectors (LISA), and is currently being implemented in \texttt{lalinference\_mcmc} 
\carl{[CITE TYSON, PROTO-TYSON]}.
